Longitudinal Data Cleaning - A Case of the NLSY79 Data

### 2021-02-22 report/report.Rmd
Add part of IDA.

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(brolgar)
library(patchwork)
library(kableExtra)
```

@@

```{r dob-tidy}
## tidy the date of birth data
dob_tidy <- new_data_qnames %>%
  select(CASEID_1979,
         starts_with("Q1-3_A~")) %>%
  mutate(dob_year = case_when(
                    # if the years recorded in both sets match, take 79 data
                    `Q1-3_A~Y_1979` == `Q1-3_A~Y_1981` ~ `Q1-3_A~Y_1979`,
                    # if the year in the 81 set is missing, take the 79 data
                    is.na(`Q1-3_A~Y_1981`) ~ `Q1-3_A~Y_1979`,
                    # if the sets don't match for dob year, take the 79 data
                    `Q1-3_A~Y_1979` != `Q1-3_A~Y_1981` ~ `Q1-3_A~Y_1979`),
        dob_month = case_when(
                    # if months recorded in both sets match, take 79 data
                    `Q1-3_A~M_1979` == `Q1-3_A~M_1981` ~ `Q1-3_A~M_1979`,
                    # if month in 81 set is missing, take the 79 data
                    is.na(`Q1-3_A~M_1981`) ~ `Q1-3_A~M_1979`,
                    # if sets don't match for dob month, take the 79 data
                    `Q1-3_A~M_1979` != `Q1-3_A~M_1981` ~ `Q1-3_A~M_1979`),
        # flag if there is a conflict between dob recorded in 79 and 81
        dob_conflict = case_when(     
                      (`Q1-3_A~M_1979` != `Q1-3_A~M_1981`) & !is.na(`Q1-3_A~M_1981`)
                      ~ TRUE,
                      (`Q1-3_A~Y_1979` != `Q1-3_A~Y_1981`) & !is.na(`Q1-3_A~Y_1981`)
                      ~ TRUE,
                      (`Q1-3_A~Y_1979` == `Q1-3_A~Y_1981`) & 
                      (`Q1-3_A~M_1979` == `Q1-3_A~M_1981`) ~ FALSE,
                      is.na(`Q1-3_A~M_1981`) | is.na(`Q1-3_A~Y_1981`) ~ FALSE)) %>%
  select(CASEID_1979,
         dob_month,
         dob_year,
         dob_conflict)
```

@@

```{r demog-tidy}
## tidy the gender and race variables
demog_tidy <- categories_qnames %>%
  select(CASEID_1979,
         SAMPLE_RACE_78SCRN,
         SAMPLE_SEX_1979) %>%
  rename(gender = SAMPLE_SEX_1979,
         race = SAMPLE_RACE_78SCRN)
```

@@

```{r tidy-grade}
# tidy the grade 
demog_education <- new_data_qnames %>%
  as_tibble() %>%
  # in 2018, the variable's name is Q3-4_2018, instead of HGC_2018
  rename(HGC_2018 = `Q3-4_2018`) %>%
  select(CASEID_1979,
         starts_with("HGCREV"),
         "HGC_2012",
         "HGC_2014",
         "HGC_2016",
         "HGC_2018") %>%
  pivot_longer(!CASEID_1979,
               names_to = "var",
               values_to = "grade") %>%
  separate("var", c("var", "year"), sep = -4) %>%
  filter(!is.na(grade)) %>%
  select(-var)
```

@@

## getting the highest year of completed education ever
highest_year <- demog_education %>%
  group_by(CASEID_1979) %>%
  mutate(hgc_i = max(grade)) %>%
  filter(hgc_i == grade) %>%
  filter(year == first(year)) %>%
  rename(yr_hgc = year) %>%
  select(CASEID_1979, yr_hgc, hgc_i) %>%
  ungroup() %>%
  mutate('hgc' = ifelse(hgc_i == 0, "NONE", ifelse(hgc_i == 1, "1ST GRADE",
                 ifelse(hgc_i == 2, "2ND GRADE", ifelse(hgc_i == 3, "3RD GRADE",
                 ifelse(hgc_i >= 4 & hgc_i <= 12, paste0(hgc_i,"TH GRADE"),
                 ifelse(hgc_i == 13, "1ST YEAR COL",
                 ifelse(hgc_i == 14, "2ND YEAR COL",
                 ifelse(hgc_i == 15, "3RD YEAR COL",
                 ifelse(hgc_i == 95, "UNGRADED", 
                        paste0((hgc_i - 12), "TH YEAR COL")))))))))))
```

@@

```{r tidy-hours}
# make a list for years where we used the "QES-52A"
year_A <- c(1979:1987, 1993)
#function to get the hour of work
get_hour <- function(year){
  if(year %in% year_A){
   temp <- new_data_qnames %>%
    select(CASEID_1979,
            starts_with("QES-52A") &
              ends_with(as.character(year)))} 
  else{
    temp <- new_data_qnames %>%
    select(CASEID_1979,
            starts_with("QES-52D") &
              ends_with(as.character(year)))} 
    temp %>% 
      pivot_longer(!CASEID_1979,
                 names_to = "job",
                 values_to = "hours_work") %>%
      separate("job", c("job", "year"), sep = -4) %>%
      mutate(job = paste0("job_", substr(job, 9, 10))) %>%
      rename(id = CASEID_1979)
}

@@

```{r tidy-rate}
get_rate <- function(year) {
  new_data_qnames %>%
     select(CASEID_1979,
            starts_with("HRP") &
              ends_with(as.character(year))) %>%
    pivot_longer(!CASEID_1979, names_to = "job", values_to = "rate_per_hour") %>%
    separate("job", c("job", "year"), sep = -4) %>%
    mutate(job = paste0("job_0", substr(job, 4, 4))) %>%
    rename(id = CASEID_1979)
}
rates <- list()
for(ayear in c(1979:1994, 1996, 1998, 2000, 2002, 2004, 2006, 2008, 2010, 
               2012, 2014, 2016, 2018)) {
   rates[[ayear]] <- get_rate(ayear)
}
rates_all <- bind_rows(!!!rates)
# join hours and rates variable
hours_wages <- left_join(rates_all, 
                         hours_all, 
                         by = c("id", "year", "job")) %>%
  # set the 0 value in rate_per_hour as NA 
  mutate(rate_per_hour = ifelse(rate_per_hour == 0, NA,
                                rate_per_hour))
head(hours_wages)
```

@@

## Robust Linear Model for Influential Values Treatment

# Conclusion

>>>>

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(brolgar)
library(patchwork)
library(kableExtra)
library(MASS)
```

@@

```{r dob-tidy}
## tidy the date of birth data
dob_tidy <- new_data_qnames %>%
  dplyr::select(CASEID_1979,
         starts_with("Q1-3_A~")) %>%
  mutate(dob_year = case_when(
                    # if the years recorded in both sets match, take 79 data
                    `Q1-3_A~Y_1979` == `Q1-3_A~Y_1981` ~ `Q1-3_A~Y_1979`,
                    # if the year in the 81 set is missing, take the 79 data
                    is.na(`Q1-3_A~Y_1981`) ~ `Q1-3_A~Y_1979`,
                    # if the sets don't match for dob year, take the 79 data
                    `Q1-3_A~Y_1979` != `Q1-3_A~Y_1981` ~ `Q1-3_A~Y_1979`),
        dob_month = case_when(
                    # if months recorded in both sets match, take 79 data
                    `Q1-3_A~M_1979` == `Q1-3_A~M_1981` ~ `Q1-3_A~M_1979`,
                    # if month in 81 set is missing, take the 79 data
                    is.na(`Q1-3_A~M_1981`) ~ `Q1-3_A~M_1979`,
                    # if sets don't match for dob month, take the 79 data
                    `Q1-3_A~M_1979` != `Q1-3_A~M_1981` ~ `Q1-3_A~M_1979`),
        # flag if there is a conflict between dob recorded in 79 and 81
        dob_conflict = case_when(     
                      (`Q1-3_A~M_1979` != `Q1-3_A~M_1981`) & !is.na(`Q1-3_A~M_1981`)
                      ~ TRUE,
                      (`Q1-3_A~Y_1979` != `Q1-3_A~Y_1981`) & !is.na(`Q1-3_A~Y_1981`)
                      ~ TRUE,
                      (`Q1-3_A~Y_1979` == `Q1-3_A~Y_1981`) & 
                      (`Q1-3_A~M_1979` == `Q1-3_A~M_1981`) ~ FALSE,
                      is.na(`Q1-3_A~M_1981`) | is.na(`Q1-3_A~Y_1981`) ~ FALSE)) %>%
  dplyr::select(CASEID_1979,
         dob_month,
         dob_year,
         dob_conflict)
```

@@

```{r demog-tidy}
## tidy the gender and race variables
demog_tidy <- categories_qnames %>%
  dplyr::select(CASEID_1979,
         SAMPLE_RACE_78SCRN,
         SAMPLE_SEX_1979) %>%
  rename(gender = SAMPLE_SEX_1979,
         race = SAMPLE_RACE_78SCRN)
```

@@

```{r tidy-grade}
# tidy the grade 
demog_education <- new_data_qnames %>%
  as_tibble() %>%
  # in 2018, the variable's name is Q3-4_2018, instead of HGC_2018
  rename(HGC_2018 = `Q3-4_2018`) %>%
  dplyr::select(CASEID_1979,
         starts_with("HGCREV"),
         "HGC_2012",
         "HGC_2014",
         "HGC_2016",
         "HGC_2018") %>%
  pivot_longer(!CASEID_1979,
               names_to = "var",
               values_to = "grade") %>%
  separate("var", c("var", "year"), sep = -4) %>%
  filter(!is.na(grade)) %>%
  dplyr::select(-var)
```

@@

## getting the highest year of completed education ever
highest_year <- demog_education %>%
  group_by(CASEID_1979) %>%
  mutate(hgc_i = max(grade)) %>%
  filter(hgc_i == grade) %>%
  filter(year == first(year)) %>%
  rename(yr_hgc = year) %>%
  dplyr::select(CASEID_1979, yr_hgc, hgc_i) %>%
  ungroup() %>%
  mutate('hgc' = ifelse(hgc_i == 0, "NONE", ifelse(hgc_i == 1, "1ST GRADE",
                 ifelse(hgc_i == 2, "2ND GRADE", ifelse(hgc_i == 3, "3RD GRADE",
                 ifelse(hgc_i >= 4 & hgc_i <= 12, paste0(hgc_i,"TH GRADE"),
                 ifelse(hgc_i == 13, "1ST YEAR COL",
                 ifelse(hgc_i == 14, "2ND YEAR COL",
                 ifelse(hgc_i == 15, "3RD YEAR COL",
                 ifelse(hgc_i == 95, "UNGRADED", 
                        paste0((hgc_i - 12), "TH YEAR COL")))))))))))
```

@@

```{r tidy-hours}
# make a list for years where we used the "QES-52A"
year_A <- c(1979:1987, 1993)
#function to get the hour of work
get_hour <- function(year){
  if(year %in% year_A){
   temp <- new_data_qnames %>%
    dplyr::select(CASEID_1979,
            starts_with("QES-52A") &
              ends_with(as.character(year)))} 
  else{
    temp <- new_data_qnames %>%
    dplyr::select(CASEID_1979,
            starts_with("QES-52D") &
              ends_with(as.character(year)))} 
    temp %>% 
      pivot_longer(!CASEID_1979,
                 names_to = "job",
                 values_to = "hours_work") %>%
      separate("job", c("job", "year"), sep = -4) %>%
      mutate(job = paste0("job_", substr(job, 9, 10))) %>%
      rename(id = CASEID_1979)
}

@@

```{r tidy-rate}
get_rate <- function(year) {
  new_data_qnames %>%
    dplyr::select(CASEID_1979,
            starts_with("HRP") &
              ends_with(as.character(year))) %>%
    pivot_longer(!CASEID_1979, names_to = "job", values_to = "rate_per_hour") %>%
    separate("job", c("job", "year"), sep = -4) %>%
    mutate(job = paste0("job_0", substr(job, 4, 4))) %>%
    rename(id = CASEID_1979)
}
rates <- list()
for(ayear in c(1979:1994, 1996, 1998, 2000, 2002, 2004, 2006, 2008, 2010, 
               2012, 2014, 2016, 2018)) {
   rates[[ayear]] <- get_rate(ayear)
}
rates_all <- bind_rows(!!!rates)
# join hours and rates variable
hours_wages <- left_join(rates_all, 
                         hours_all, 
                         by = c("id", "year", "job")) %>%
  # set the 0 value in rate_per_hour as NA 
  mutate(rate_per_hour = ifelse(rate_per_hour == 0, NA,
                                rate_per_hour))
head(hours_wages)
```

@@

According to @HuebnerMariannePhD2016Asat, Initial Data Analysis (IDA) is the step of inspecting and screening the data after being collected to ensure that the data is clean, valid, and ready to be deployed in the later formal statistical analysis. Moreover, @Chatfield1985TIEo argued that the two main objectives of IDA is data description, which is to assess the structure and the quality of the data; and model formulation without any formal statistical inference.

In this paper, we conducted an IDA or a preliminary data analysis to assess the consistency of the data with the cohort information that is provided by the NLSY. In addition, we also aimed to find the anomaly in the wages values using this approach. We mainly used graphical summary to do the IDA using `ggplot2`[@ggplot2] and `brolgar` [@brolgar].

As stated previously, the respondents' ages ranged from 12 to 22 when first interviewed in 1979. Hence, we would like to validate whether all of the respondents were in this range. For the demographic data validation, we used the `full_demographic` i.e. the data with the survey years 1979 sample.

```{r valid-age, echo = FALSE, fig.cap="Subjects' Age Distribution"}
age <- yowie::demographic_nlsy79 %>%
  ggplot(aes(x = age_1979)) +
  geom_bar(fill = "#45B39D")

gender <- yowie::demographic_nlsy79 %>%
  ggplot(aes(x = gender,
             fill = gender)) +
  geom_bar() +
  geom_text(aes(label = ..count..), stat = "count", vjust = 1.5, colour = "white")+
  scale_fill_manual("gender", values = c("MALE" = "#2471A3", "FEMALE" = "#9B59B6"))+
  theme(legend.position = "none")

race <- yowie::demographic_nlsy79 %>%
  ggplot(aes(x = race,
             fill = race)) +
  geom_bar() +
  geom_text(aes(label = ..count..), stat = "count", vjust = 1.5, colour = "white") +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 10))

age / (gender|race)
```

### Robust Linear Model for Influential Values Treatment

# Summary
